---
title: "Target aggregate data adjustment transportability analysis"
author: "Core Clinical Sciences"
output: rmarkdown::html_vignette
bibliography: references.bib

vignette: >
  %\VignetteIndexEntry{Target aggregate data adjustment transportability analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE}
library(TransportHealth)

expit <- function(x) 1/(1+exp(-x))

# ---------- Generate test data for TADA for TransportHealth ----------
generateTestDataTADA <- function() {
  
  # ---------- Generate study data in IPD ----------
  nStudy <- 1000
  
  sexStudy <- rbinom(nStudy, 1, 0.5) # Male is 1, so female is baseline
  stressStudy <- rbinom(nStudy, 1, 0.4) # Stressed is 1
  med2Study <- rbinom(nStudy, 1, 0.1) # 1 means taking other med
  percentBodyFatStudy <- rnorm(nStudy, 28 - 13 * sexStudy, 2)
  toxicGradeStudy <- sample(c("Low", "Medium", "High"), size = nStudy, replace = TRUE, prob = c(0.7, 0.15, 0.15))
  
  # treatment
  med1Study <- rbinom(nStudy, 1, expit(0.2 * sexStudy - 0.02 * percentBodyFatStudy + 0.1 * stressStudy))
  # response
  sysBloodPressureStudy <- rnorm(nStudy, 100 + 5 * sexStudy + 0.5 * percentBodyFatStudy + 5 * stressStudy -
                                   5 * med1Study + med1Study * (-5 * med2Study + 7 * stressStudy))
  
  
  # Put all variables together
  studyData <- data.frame( sysBloodPressure = sysBloodPressureStudy, # response
                           med1 = as.factor(med1Study), # treatment
                           
                           sex = as.factor(sexStudy), 
                           stress = as.factor(stressStudy), 
                           med2 = as.factor(med2Study), 
                           #toxicGrade = as.factor(toxicGradeStudy),
                           percentBodyFat = percentBodyFatStudy)
  
  
  
  # ---------- Generate target data in AgD ----------
  nTarget <- 1500
  
  sexTarget <- rbinom(nTarget, 1, 0.3) # Male is 1, so female is baseline
  stressTarget <- rbinom(nTarget, 1, 0.7) # Stressed is 1
  med2Target <- rbinom(nTarget, 1, 0.3) # 1 means taking other med
  percentBodyFatTarget <- rnorm(nTarget, 26 - 12 * sexTarget, 2)
  toxicGradeTarget <- sample(c("Low", "Medium", "High"), size = nTarget, replace = TRUE, prob = c(0.45, 0.35, 0.20))

  # Put all variables together
  targetData <- data.frame(N = nTarget,
                           
                           # binary
                           sex_count = sum(sexTarget), # count of Male
                           sex_prop = 0.93, # an incorrect value to test the replacement function, the new prop should be calculated based on 
                           
                           stress_prop = sum(stressTarget)/nTarget, # proportion of stress
                           
                           med2_PROP = sum(med2Target)/nTarget, 
                           
                           # ordinal with three subclass
                           # toxicGrade_LOW_prop = as.numeric(table(toxicGradeTarget)["Low"]) / nTarget,
                           # toxicGrade_Medium_COUNT = as.numeric(table(toxicGradeTarget)["Medium"]),
                           
                           # ===========
                           # This sentence below doesn't make sense to me. Does it mean that we don't need both proportions?
                           # What is a reference-level candidate?
                           # Are we doing categorical effect modifiers and covariates as well? That's what the OR
                           # seems to imply
                           #===========
                           
                           # # We don't need to prop for reference level candidate
                           # # like: for male and female we just need to provide one of two and it's informative
                           # # OR: user could provide all AgD for all levels and we could detect the ref level by the first element of factor IPD
                           # toxicGrade_high_prop = as.numeric(table(toxicGradeTarget)["High"]) / nTarget,
                           
                           # continuous
                           percentBodyFat_mean = mean(percentBodyFatTarget),
                           percentBodyFat_sd = sd(percentBodyFatTarget),
                           percentBodyFat_median = median(percentBodyFatTarget)
                           )
  

  return(list(studyData = studyData, 
              aggregateTargetData = targetData))
}

testData <- generateTestDataTADA()

```

# Introduction

In this vignette, we demonstrate how to use `TransportHealth` for weighting-based transportability analysis when individual participant-level data (IPD) is available for the original population, while only aggregate-level data (AgD) is available for the target sample.

## Brief introduction to Target aggregate data adjustment (TADA)

In transportability and generalizability analysis using target aggregate data adjustment (TADA), the process differs from traditional inverse odds of participation weighting (IOPW) or g-computation due to the absence of individual-level data (IPD) in the target population. Instead of using logistic regression to obtain weights that balance the distribution of effect modifiers in the source and target data, TADA uses a method of moments approach with the aggregate-level data (AgD) from the target population. 
<!--It would be good to be more clear here -- Method of moments approach to? obtain weights and balance the distribution of effect modifiers? -->
This is combined with the usual propensity score weighting approach to adjust for confounding in the source data to fit marginal structural models.

<!-- And then to fit MSMs? -->

# Example

Suppose we are interested in estimating the causal effect of a medication on systolic blood pressure in a target population, but we were only able to conduct an observational study using samples from the study population. To obtain unbiased causal effect estimates using the study sample, we account for the following covariates: sex, body fat percentage, and stress level.

We know that the effectiveness of the medication depends on two effect modifiers: 1) stress levels and; 2) whether patients are taking another medication.

*Note that the covariates adjusted for in the study data can also be effect modifiers.*

## Coded variables:

-   Medication - `med1`

    -   `1` for treated
    -   `0` for untreated

-   Systolic blood pressure (SBP) - `sysBloodPressure` (continuous)

-   Sex - `sex`

    -   `1` for male
    -   `0` for female

-   Body fat percentage - `percentBodyFat` (continuous)

-   Stress level - `stress`

    -   `1` for stressed
    -   `0` for not stressed

-   Medication 2 - `med2`

    -   `1` for treated
    -   `0` for untreated

## Analyses

First, for the implementation of TADA in `TransportHealth` specifically, the data from the study and target population should be in separate data frames in the `R` environment. The study data should contain IPS on response, treatment, covariates and effect modifiers, while the target data should contain only aggregate-level summary statistics of covariates and effect modifiers. These summary statistics should be formatted as proportions, counts, means, medians and standard deviations.

Suppose that we have the study and target data separately as follows.

```{r exampleData, echo = T}
names(testData)
print("Study data:")
head(testData$studyData)
print("Target data:")
head(testData$aggregateTargetData)

```

Here for the target data, the names of the aggregate level statistics corresponding to each variable should be formatted as follows.

-   For continuous variables, the mean of a given variable should be named `variable_mean`, the standard deviation should be named `variable_sd`, and the median should be named `variable_median`.

-   The proportion or count of a binary variable can be named `variable_prop` or `variable_count`, respectively (corresponding to the category labelled as 1 in the study data).
<!-- What do you mean by labelled as 1? Do you mean the reference? -->

-   The naming method of categorical variables (with more than two levels) is specified as `variable_category_AgD`. For example, for an ordinal variable `grade` with three categories (low, medium and high), the proportions of each category recorded in the target data should be named `grade_low_prop`, `grade_medium_prop` and `grade_high_prop`. This is just to demonstrate the naming format for categorical variables; there are no categorical variables in this example.

We can now perform transportability analysis using TADA with the `transportTADA` functions.

```{r blankFunction, echo = T, eval = F}
transportTADA(msmFormula, 
              propensityScoreModel,
              matchingCovariates,
              propensityWeights,
              participationWeights,
              treatment,
              response,
              family,
              studyData,
              aggregateTargetData)

```

### Arguments for the `transportTADA` function

This function requires specification of the following arguments:

-   `msmFormula`: A formula expressing the marginal structural model (MSM) to be fit.

-   `propensityScoreModel`: A formula or a `glm` object expressing the propensity model, i.e. a model of treatment assignment in terms of covariates.

<!-- Is this information below correct for TADA too? Why would it default to logistic when we are doing MoM?-->

If a formula is provided, logistic regression is used by default. Custom propensity weights from other weighting methods can also be provided to the `customPropensity` argument instead; in this case, do not set `propensityScoreModel` because it is `NULL` by default and will be overridden.

-   `matchingCovariates`: A vector of user-specified covariates which are used for the data matching to obtain participation weights when no custom weights provided.

-   `propensityWeights`: A vector of custom weights balancing covariates between treatments. Providing them will override the formula or model provided by `propensityScoreModel`. This vector should have as any entries as the sample size of the study data.

-   `participationWeights`: A vector of custom weights balancing effect modifiers between study and target populations. This vector should have as any entries as the sample size of the study data.

-   `treatment`: The name of the variable indicating treatment.

-   `response`: The name of the variable indicating response.

-   `family`: The type of outcome model to be fit. This can be any of the families used in `glm`, one of `"coxph"` or `"survreg"`. The `"coxph"` and `"survreg"` options are for survival analysis and will use default options of these methods from the `survival` package.

-   `studyData`: The individual participant data (IPD) of study population.

-   `aggregateTargetData`: The aggregate-level data (AgD) of target population.

### Specification of transportability analysis

These components are put together as follows.

Recall that: - `sysBloodPressure` is the response

-   `med1` is the treatment

-   `sex`, `percentBodyFat`, `toxicGrade`, and `stress` are covariates to be controlled in the original study

-   `med2` (other medication) and `stress` are effect modifiers of interest.

```{r TADA, echo = T}
result <- transportTADA(
  # MSM formula
  msmFormula = sysBloodPressure ~ med1,
  
  # Propensity model
  propensityScoreModel = med1 ~ sex + percentBodyFat + stress,
  
  # Matching covariates
  matchingCovariates = c("stress", "med2"),
  
  # Name of treatment variable
  treatment = "med1", 
  
  # Name of response variable
  response = "sysBloodPressureStudy",
  
  # Type of MSM
  family = gaussian,
  
  # Study data
  studyData = testData$studyData,
  
  # Target data
  aggregateTargetData = testData$aggregateTargetData)

```

### Producing statistical results

To show the results of the analysis, we can use the `base::summary` function, similar to how one would use the `lm` function for fitting a linear model.

The `base::summary` function prints out a table of unweighted and weighted SMDs of covariates between treatment groups, a pre-post weighting table which includes unweighted and weighted aggregate level summary as well as pre- and post- difference values of effect modifiers of interest between study data and target data, and summaries of the fitted outcome model and the final marginal structural model (MSM) fit with the correct standard errors calculated using sandwich estimator or bootstrap.
<!-- Sandwich estimator or bootstrap depending on?? -->
Note that scientific conclusions should only be drawn from the MSM.

```{r TADA-summary, echo = T}
summary(result)
```

**Positivity** and **conditional exchangeability** are two key assumptions that can affect the validity of transportability analysis.

`Positivity` is the assumption that at all observed levels of covariates and effect modifiers, the probabilities of being in the treatment group and being the study are neither 0 nor 1, respectively.

To evaluate this assumption for the treatment assignment, use the `plot` function with `type = "propensityHist"`. This outputs mirrored histogram of probabilities of being in the treatment group for different treatment groups. Non-overlap of the ranges of the histograms suggest violations of the positivity assumption [@positivity].

```{r TADA-hist, echo = T}
base::plot(result, type = "propensityHist")
```

The `plot` function with `type = "participationHist"` provides the histogram of participation weights and the effective sample size (ESS), which indicates the possible existence of extreme weights and the degree of ESS reduction. This information can justify the suitability of the TADA method.

<!-- I don't think justify is the right word. Do you mean assess? It would be helpful to know what is meant by "suitability" -- what would this plot have to tell you for TADA to not be suitable?  -->

```{r TADA-hist, echo = T}
base::plot(result, type = "participationHist")
```

`Conditional exchangeability` is roughly the assumption that the only possible confounding is due to the controlled covariates and effect modifiers. Under this assumption, TADA estimates will be reliable if the weighted distributions of covariates and effect modifiers are similar between treatment groups and the study data, respectively. This can be (partially) evaluated using standardized mean differences (SMDs), which are shown in table form by the `summary` function. The `plot` function with `type = "propensitySMD"` provides graphical versions of this table. A general guideline is that an SMD of below 0.1 indicates balance, but this threshold is arbitrary and left to the choice of the analyst [@balance].

```{r TADA-smd, echo = T}
base::plot(result, type = "propensitySMD")

```

Model coefficient plots showing confidence intervals of the effect estimates are provided by `plot` function with `type = "msm"`. The standard errors are the correct ones calculated by sandwich estimator or bootstrap.
<!-- Same as above -- "depending on...." -->

```{r TADA-msm, echo = T}
base::plot(result, type = "msm")
```

Note that all plot outputs are designed to be generated with `ggplot2`. They can be customized further.

## References

::: {#refs}
:::
