---
title: "Transportability analysis with interpolated g-computation"
output: rmarkdown::html_vignette
bibliography: references.bib

vignette: >
  %\VignetteIndexEntry{Transportability analysis with interpolated g-computation}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = F}
library(TransportHealth)
set.seed(20240501)

expit <- function(x) 1/(1+exp(-x))

# Generate test data for TransportHealth. This data is not meant to be officially put in the package. It is simply to test whether the implemented functions in the package accept the correct inputs and outputs.
generateTestData <- function() {
  # Generate study data
  nStudy <- 1000
  sexStudy <- rbinom(nStudy, 1, 0.5) # Male is 1, meaning that female is baseline (0)
  stressStudy <- rbinom(nStudy, 1, 0.4) # Stressed is 1, not stressed is 0
  med2Study <- rbinom(nStudy, 1, 0.1) # 1 means taking a medication other than the one under investigation, 0 means not taking another medication
  percentBodyFatStudy <- rnorm(nStudy, 28 - 13 * sexStudy, 2) #this variable is the percent of body fat each participant has
  med1Study <- rbinom(nStudy, 1, expit(0.2 * sexStudy - 0.02 * percentBodyFatStudy + 0.1 * stressStudy)) #this variable is the main intervention or treatment being studied. 1 means the participant is receiving the medication under investigation, 0 means they are not.
  sysBloodPressureStudy <- rnorm(nStudy, 100 + 5 * sexStudy + 0.5 * percentBodyFatStudy + 5 * stressStudy -
                                   5 * med1Study + med1Study * (-5 * med2Study + 7 * stressStudy)) #this variable is the main outcome. It is a continuous measure of systolic blood pressure
  
  # Put all variables together
  studyData <- data.frame(sysBloodPressure = sysBloodPressureStudy, med1 = as.factor(med1Study), sex = as.factor(sexStudy), stress = as.factor(stressStudy), med2 = as.factor(med2Study), percentBodyFat = percentBodyFatStudy)
  
  # Generate target data
  nTarget <- 1500
  sexTarget <- rbinom(nTarget, 1, 0.3) # Male is 1, so female is baseline (0)
  stressTarget <- rbinom(nTarget, 1, 0.7) # Stressed is 1, so 0 is not stressed
  med2Target <- rbinom(nTarget, 1, 0.3) # 1 means taking a medication other than the one under investigation, 0 means not taking another medication
  percentBodyFatTarget <- rnorm(nTarget, 26 - 12 * sexTarget, 2) #this variable is the percent of body fat each participant has
  
  # Put all variables together
  targetData <- data.frame(sex = as.factor(sexTarget), stress = as.factor(stressTarget), med2 = as.factor(med2Target), percentBodyFat = percentBodyFatTarget)
  
  return(list(studyData = studyData, targetData = targetData))
}

generateTestDataInterpolated <- function() {
  testData <- generateTestData()
  studyData <- testData$studyData
  targetData <- testData$targetData
  studyData$percentBodyFatDicho <- as.numeric(studyData$percentBodyFat >= 17)
  targetData$percentBodyFatDicho <- as.numeric(targetData$percentBodyFat >= 17)
  
  nStudy <- nrow(studyData)
  
  mainTreatmentEffect <- with(studyData, mean(sysBloodPressure[med1 == 1]) - mean(sysBloodPressure[med1 == 0]))
  mainSE <- with(studyData, sqrt((var(sysBloodPressure[med1 == 1]) * (length(which(med1 == 1)) - 1) +
                            var(sysBloodPressure[med1 == 0]) * (length(which(med1 == 0)) - 1)) / (nStudy - 2)) * 
                            sqrt((1/length(which(med1 == 1))) + (1/length(which(med1 == 0)))))
  
  
  effectModifiers <- c("med2", "percentBodyFatDicho")
  
  aggregateStudyData <- apply(studyData[, effectModifiers], 2, function(x) {x |> as.numeric() |> mean()})
  
  subgroupLevels <- rep(c(1,0), times = length(effectModifiers))
  
  subgroupTreatmentEffects <- sapply(1:length(subgroupLevels), function(i) {
                              with(studyData[studyData[[effectModifiers[ceiling(i/2)]]] == subgroupLevels[i],],
                                   mean(sysBloodPressure[med1 == 1]) - mean(sysBloodPressure[med1 == 0])) 
                              })
  
  subgroupSEs <- sapply(1:length(subgroupLevels), function(i) {
    with(studyData[studyData[[effectModifiers[ceiling(i/2)]]] == subgroupLevels[i],],
         sqrt((var(sysBloodPressure[med1 == 1]) * (length(which(med1 == 1)) - 1) +
              var(sysBloodPressure[med1 == 0]) * (length(which(med1 == 0)) - 1)) / (nStudy - 2)) * 
           sqrt((1/length(which(med1 == 1))) + (1/length(which(med1 == 0))))
    )
  })
  
  targetData$n <- nrow(targetData)
  
  aggregateTargetData <- apply(targetData, 2, function(x) mean(as.numeric(x)))
  names(aggregateTargetData) <- names(targetData)
  
  return(list(mainTreatmentEffect = mainTreatmentEffect,
              mainSE = mainSE,
              effectModifiers = effectModifiers,
              subgroupTreatmentEffects = subgroupTreatmentEffects,
              subgroupSEs = subgroupSEs,
              nStudy = nStudy,
              aggregateStudyData = aggregateStudyData,
              targetData = targetData,
              aggregateTargetData = aggregateTargetData))
}

testData <- generateTestDataInterpolated()
```

# Introduction

In this vignette, we demonstrate how to use `TransportHealth` for interpolated g-computation, a transportability analysis method based on network meta-interpolation [@harari] for use when only aggregate data is available for the original study sample.

## Brief introduction to interpolated g-computation

In transportability and generalizability analysis, interpolated g-computation proceeds in largely the same way as network meta-interpolation. First, a model of the treatment effect is fitted in terms of dichotomized effect modifiers. Missing effect modifier information is imputed using the best linear unbiased predictor (BLUP) in terms of observed effect modifier information. Then, an estimate of the treatment effect in the target population is calculated with the fitted model using the summary information of effect modifiers in the target sample.

# Example

Suppose we are interested in estimating the causal effect of a medication on systolic blood pressure in a target population, but we were only able to conduct a randomized clinical trial using samples from the a different population.

We know that the effectiveness of the medication depends on two effect modifiers: 1) stress level, and 2) whether patients are taking another medication.

## Coded variables:

-   Medication - `med1`

    -   `1` for treated
    -   `0` for untreated

-   Systolic blood pressure (SBP) - `sysBloodPressure` (continuous)

-   Sex - `sex`

    -   `1` for male
    -   `0` for female

-   Body fat percentage - `percentBodyFat` (continuous)

-   Stress level - `stress`

    -   `1` for stressed
    -   `0` for normal

-   Medication 2 - `med2`

    -   `1` for treated
    -   `0` for untreated

## Analyses

First, for the implementation of interpolated g-computation in `TransportHealth` specifically, the study data should be aggregate-level. In particular, effect modifiers should be dichotomized. For this case study, body fat percentage has been dichotomized with 17% as the threshold. For clarity, we assume that dichotomized variables are coded as 0-1 with 0 being the baseline. The aggregate-level study data should have:

-   The estimated treatment effect and its estimated standard error
-   The estimated subgroup treatment effect within both (marginal) levels of each effect modifier and its estimated standard error
-   Proportion of 1s for each effect modifier in the study sample (summary study data)
-   The sample size of the original study

The subgroup treatment effects, standard errors and summary study data should be provided in a vector in the same order. Subgroup effects and standard errors corresponding to the levels of the same effect modifier should be next to each other with the subgroup effect of the 1-level placed first. Also, provide a vector which specifies the order in which the subgroup treatment effects and summary study data are provided. In our study, the aggregate-level study data is formatted as follows.

```{r exampleStudyData, echo = T}
print("Treatment effect:")
print(testData$mainTreatmentEffect)
print("Standard error of treatment effect:")
print(testData$mainSE)
print("Effect modifier names:")
print(testData$effectModifiers)
print("Subgroup effects:")
print(testData$subgroupTreatmentEffects)
print("Standard errors of subgroup effects:")
print(testData$subgroupSEs)
print("Summary study data:")
print(testData$aggregateStudyData)
print("Sample size:")
print(testData$nStudy)
```

Since `testData$effectModifiers` is `c(med2, percentBodyFatDicho)`, the subgroup effects provided in the `testData$subgroupTreatmentEffects` are, in order, those corresponding to the 1 group and the 0 group of `med2`, and then those corresponding to the 1 group and 0 group of `percentBodyFatDicho`. This ordering is the same for `testData$subgroupSEs`. Likewise, `testData$aggregateStudyData` provides proportions of 1s for `med2` and `percentBodyFatDicho`, in this order. The ordering is important for the modelling function to process the data correctly.

On the other hand, the target data may be individual patient-level data or aggregate data. For the former, effect modifiers should still be dichotomized in the target data. For the latter, provide the proportion of 1s for each dichotomized effect modifier, along with the sample size, in a named vector; the ordering does not matter as long as the vector is named appropriately. It is important that effect modifiers are dichotomized the same way in both the study and the target data. You may optionally provide a correlation (not variance) matrix of the dichotomized effect modifiers. If a correlation matrix is not provided, the correlation matrix is calculated from the target data when individual patient-level data is provided, and an independent correlation structure is assumed when aggregate data is provided. In this example, the target data is formatted as follows.

```{r exampleData, echo = T}
print("Target data:")
head(testData$targetData)
```

We can now perform transportability analysis using interpolated g-computation with the `transportInterpolated` function.

```{r blankFunction, echo = T, eval = F}
transportInterpolated(link,
                      effectModifiers,
                      mainTreatmentEffect,
                      mainSE,
                      subgroupTreatmentEffects,
                      subgroupSEs,
                      corrStructure = NULL,
                      studySampleSize,
                      aggregateStudyData,
                      targetData)
```

### Arguments for the `transportInterpolated` functions

The `transportInterpolated` function requires the following arguments:

-   `link`: The link function of the outcome in terms of the treatment used in the source study. If mean differences are provided, use `"identity"`. If odds ratios or hazard ratios are provided, use `"log"`.

-   `effectModifiers`: Vector of names of effect modifiers to adjust for

-   `mainTreatmentEffect`, `mainSE`: The estimated treatment effect in the original study and its estimated standard error

-   `subgroupTreatmentEffects`, `subgroupSEs`: The vector of estimated subgroup treatment effects for each (marginal) level of each effect modifier and their estimated standard errors, formatted as specified above

-   `corrStructure`: The correlation matrix of effect modifiers. This is optional to provide if IPD target data is provided, and defaults to an independent structure if aggregate-level target data is provided.

-   `studySampleSize`: Sample size of original study data

-   `aggregateStudyData`: Summary data of original study data

-   `targetData`: IPD or aggregate-level target data.

### Specification of transportability analysis

Recall that: - `sysBloodPressure` is the response

-   `med1` is the treatment

-   `med2` (other medication) and `stress` are effect modifiers of interest.

We supply arguments to the `transportInterpolated` function as follows. As `sysBloodPressure` is a continuous outcome, the effect estimates are mean differences, so `link = "identity"` will be used.

```{r PreparedModel, echo = T}
result <- transportInterpolated(link = "identity",
                                effectModifiers = testData$effectModifiers,
                                mainTreatmentEffect = testData$mainTreatmentEffect,
                                mainSE = testData$mainSE,
                                subgroupTreatmentEffects = testData$subgroupTreatmentEffects,
                                subgroupSEs = testData$subgroupSEs,
                                studySampleSize = testData$nStudy,
                                aggregateStudyData = testData$aggregateStudyData,
                                targetData = testData$targetData)
```

### Producing statistical results

To show the results of the analysis, use `summary` like you would for `lm` when fitting a linear model. Using `summary` will print out the transported effect estimate and its estimated standard error, the link function, and summaries of the data provided to the function. Note that scientific conclusions should only be drawn from this output.

```{r GComputation-summary, echo = T}
summary(result)
```

We have found via simulation that interpolated g-computation tends to perform less well when the degree of effect modification is larger than the main treatment effect, so conclusions from `transportInterpolated` in such situations should be made with caution. The magnitude of effect modification can be informally evaluated by looking at forest plots of the original study from which the treatment effects are retrieved, or by looking at the `summary` output of `transportInterpolated`. More specifically, the output contains the subgroup effects provided by the user. Using either method, one should compare the difference in subgroup effects between levels of the same effect modifier with the magnitude of the main treatment effect to evaluate whether the interaction effect is larger than the main effect.

To obtain a coefficient plot of estimates, use the `plot` function.

```{r GComputation-plot, echo = T}
plot(result)
```

Like other methods supported by `TransportHealth`, the validity of interpolated g-computation depends on untestable causal inference assumptions, including stable unit treatment value (SUTVA), conditional exchangeability, positivity and consistency [@ling-review; @degtiar-review]. However, interpolated g-computation does not have readily available diagnostics to evaluate if these assumptions are likely to hold or not. Due to the limited amount of information from the original study, causal inference assumptions should be evaluated using contextual knowledge about study design.

Additionally, it is expected that interpolated g-computation will perform less well than inverse odds of participation weights and g-computation, as a lot of information is lost when only aggregate-level data is available for the original study data. As a result, we suggest that only one effect modifier is adjusted for when using interpolated g-computation.

## References

::: {#refs}
:::
